\pkg{sparseHessianFD}: An R Package for Estimating Sparse Hessian Matrices

### 2016-12-24 nobuild/jss_revision/sparseHessianFD.Rnw
refining the revision

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  Like
\pkg{sparseHessianFD} it includes functions to compute Hessians from
user-supplied gradients (through the \func{jacobian} function), and
implements both the finite differencing and complex step methods.  Its
most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
\pkg{sparseHessianFD}.  All of the subsequent speed and scalability
results were generated on an Apple Mac Pro with a 2.7 GHz
12-core Intel Xeon E5 processor and 64 GB of RAM, running MacOS
version 10.12.2, and \proglang{R} version 3.3.2.

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  As in the previous
section, the total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 20 replications) for the time (in
milliseconds) to compute a Hessian using each package.  The difference in run times is dramatic,
especially because the computation time for \pkg{numDeriv} grows
quadratically in the number of variables. Times
were collected using the \pkg{microbenchmark} package
\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
is available as an online supplement to this paper, and in the
\func{doc/} directory of the installed package.

@@

To help us understand just how scalable
\pkg{sparseHessianFD} is, we ran another set of simulations, for the
same hierarchical model, for different values of $N$ and $k$.  We then
recorded the run times for different steps in the sparse Hessian
estimation, across 200 replications.  The steps are summarized in Table~\ref{tab:timeMeasures}.

@@

In the plots in Figure~\ref{fig:timing}, the number of heterogeneous units ($N$) is
on the x-axis, and median run time, in milliseconds, is on the y-axis.  Each panel shows the
relationship between $N$ and run time for a different measure of time,
and each curve in a panel represents a different number of within-unit
parameters ($k$).

>>>>

As far as we know, \pkg{numDeriv} \citep{R_numDeriv} is the only other \proglang{R} package
that computes numerical approximations to derivatives.  Like
\pkg{sparseHessianFD} it includes functions to compute Hessians from
user-supplied gradients (through the \func{jacobian} function), and
implements both the finite differencing and complex step methods.  Its
most important distinction from \pkg{sparseHessianFD} is that it treats all Hessians as dense.  Thus, we
will use  \pkg{numDeriv} as the baseline against which we can compare the performance of
\pkg{sparseHessianFD}.

To prepare Table~\ref{tab:numDeriv}, we
estimated Hessians of the log posterior density in Equation~\ref{eq:LPD} with different numbers of heterogeneous
units ($N$) and within-unit parameters ($k$).  The total number of
variables is $M=(N+1)k$.  Table~\ref{tab:numDeriv} shows the mean
and standard deviations (across 500 replications) for the time (in
milliseconds) to compute a Hessian using functions for both the finite
difference and complex step methods from each package. Times were generated on a
compute node running Scientific Linux 6 (64-bit) with an 8-core Intel
Xeon X5560 processor (2.80 GHz) with 24 GB of RAM, and collected using the \pkg{microbenchmark} package
\citep{R_microbenchmark}. Code to replicate Table~\ref{tab:numDeriv}
is available as an online supplement to this paper, and in the
\func{doc/} directory of the installed package. In
Table~\ref{tab:numDeriv} we see that computation times using
\pkg{sparseHessianFD} and considerably shorter than those using \pkg{numDeriv}.

@@

To help us understand just how scalable
\pkg{sparseHessianFD} is, we ran another set of simulations, for the
same hierarchical model, for different values of $N$ and $k$.  We then
recorded the run times for different steps in the sparse Hessian
estimation, across 200 replications.  The steps are summarized in
Table~\ref{tab:timeMeasures}.  The times were generated on an
Apple Mac Pro with a 12-core Intel Xeon E5-2697 processor (2.7 GHz)
with 64 GB of RAM.

@@

In the plots in Figure~\ref{fig:timing}, the number of heterogeneous units ($N$) is
on the x-axis, and median run time, in milliseconds, is on the y-axis.  Each panel shows the
relationship between $N$ and run time for a different step in the algorithm,
and each curve in a panel represents a different number of within-unit
parameters ($k$).
